Streaking and Clutching in Round-Based First Person Shooter (FPS) Esports

### 2022-07-15 presentation.qmd
better pres name, start looking at b2b

---
title: "Momentum in Round-Based First-Person-Shooter (FPS) E-Sports"
author: "Tony ElHabr"
format:
  revealjs:
    incremental: false
    theme: [dark, theme.scss]
execute: 
  include: false
  echo: false
  eval: false
---

@@

heatmap_tile <- function(..., .fill = '') {
  list(
    ...,
    geom_rect(
      color = blackish_background,
      aes(
        fill = .data[[.fill]],
        xmin = pre_cumu_w, 
        ymin = pre_cumu_l,
        xmax = pre_cumu_w + 1,
        ymax = pre_cumu_l + 1
      )
    )
  )
}

@@

p_o_round_win_prop <- o_win_prop |> 
  ggplot() +
  common_heatmap_layers() +
  heatmap_tile(.fill = 'win_round_prop') +
  add_text_layers(.f = add_o_round_win_text_layer) +
  ggsci::scale_fill_material('teal') +
  labs(
    title = 'Offensive Round Win %'
  )
p_o_round_win_prop
save_heatmap(p_o_round_win_prop, width = 7.5, height = 5)
```

@@

max_diff_win_round_prop <- max(abs(o_win_prop$diff_win_round_prop))
p_o_round_win_prop_over_expected <- o_win_prop |> 
  ggplot() +
  common_heatmap_layers() +
  heatmap_tile(.fill = 'diff_win_round_prop') +
  add_text_layers(.f = add_o_round_win_over_expected_text_layer) +
  scale_fill_gradient2(
    low = extract_ggsci_mid_color('purple'),
    high = extract_ggsci_mid_color('green'),
    mid = 'white',
    midpoint = 0,
    limits = c(
      -max_diff_win_round_prop, 
      +max_diff_win_round_prop
    )
  ) +
  labs(
    title = 'Offensive Round Win % Over Expected'
  )
p_o_round_win_prop_over_expected
save_heatmap(p_o_round_win_prop_over_expected, width = 7.5, height = 5)
```

@@

p_o_series_win_prop <- o_win_prop |> 
  ggplot() +
  common_heatmap_layers() +
  heatmap_tile(.fill = 'win_series_prop') +
  add_text_layers(.f = add_o_series_win_text_layer) +
  ggsci::scale_fill_material('deep-orange') +
  labs(
    title = 'Offensive Series Win %'
  )
p_o_series_win_prop
save_heatmap(p_o_series_win_prop, width = 7.5, height = 5)
```

@@

max_diff_win_series_prop <- max(abs(o_win_prop$diff_win_series_prop))
p_o_series_win_prop_over_expected <- o_win_prop |> 
  ggplot() +
  common_heatmap_layers() +
  heatmap_tile(.fill = 'diff_win_series_prop') +
  add_text_layers(.f = add_o_series_win_over_expected_text_layer) +
  scale_fill_gradient2(
    low = extract_ggsci_mid_color('brown'),
    high = extract_ggsci_mid_color('cyan'),
    mid = 'white',
    midpoint = 0,
    limits = c(
      -max_diff_win_series_prop, 
      +max_diff_win_series_prop
    )
  ) +
  labs(
    title = 'Offensive Series Win % Over Expected'
  )
p_o_series_win_prop_over_expected
save_heatmap(p_o_series_win_prop_over_expected, width = 7.5, height = 5)
```

@@

@@

@@

::: notes
-   plant at most prevalent/winningest site? plant at other site? go for kills?
:::

>>>>

---
title: "Streaking and Clutching in Round-Based First Person Shooter (FPS) Esports"
author: "Tony ElHabr"
format:
  revealjs:
    incremental: false
    theme: [dark, theme.scss]
execute: 
  include: false
  echo: false
  eval: false
---

@@

heatmap_tile <- function(..., .xmin, .xmax, .ymin, .ymax, .fill) {
  list(
    ...,
    geom_rect(
      color = blackish_background,
      aes(
        fill = .data[[.fill]],
        xmin = .data[[.xmin]], 
        ymin = .data[[.ymin]],
        xmax = .data[[.xmin]], + 1,
        ymax = .data[[.ymin]] + 1
      )
    )
  )
}

pre_cumu_heatmap_tile <- function(..., .fill) {
  heatmap_tile(
    .xmin = 'pre_cumu_w',
    .ymin = 'pre_cumu_l',
    .fill = .fill,
    ...
  )
}

@@

p_o_round_win_prop <- o_win_prop |> 
  ggplot() +
  common_heatmap_layers() +
  pre_cumu_heatmap_tile(.fill = 'win_round_prop') +
  add_text_layers(.f = add_o_round_win_text_layer) +
  ggsci::scale_fill_material('teal') +
  labs(
    title = 'Offensive Round Win %'
  )
p_o_round_win_prop
save_heatmap(p_o_round_win_prop, width = 7.5, height = 5)
```

@@

max_diff_win_round_prop <- max(abs(o_win_prop$diff_win_round_prop))
p_o_round_win_prop_over_expected <- o_win_prop |> 
  ggplot() +
  common_heatmap_layers() +
  pre_cumu_heatmap_tile(.fill = 'diff_win_round_prop') +
  add_text_layers(.f = add_o_round_win_over_expected_text_layer) +
  scale_fill_gradient2(
    low = extract_ggsci_mid_color('purple'),
    high = extract_ggsci_mid_color('green'),
    mid = 'white',
    midpoint = 0,
    limits = c(
      -max_diff_win_round_prop, 
      +max_diff_win_round_prop
    )
  ) +
  labs(
    title = 'Offensive Round Win % Over Expected'
  )
p_o_round_win_prop_over_expected
save_heatmap(p_o_round_win_prop_over_expected, width = 7.5, height = 5)
```

@@

p_o_series_win_prop <- o_win_prop |> 
  ggplot() +
  common_heatmap_layers() +
  pre_cumu_heatmap_tile(.fill = 'win_series_prop') +
  add_text_layers(.f = add_o_series_win_text_layer) +
  ggsci::scale_fill_material('deep-orange') +
  labs(
    title = 'Offensive Series Win %'
  )
p_o_series_win_prop
save_heatmap(p_o_series_win_prop, width = 7.5, height = 5)
```

@@

max_diff_win_series_prop <- max(abs(o_win_prop$diff_win_series_prop))
p_o_series_win_prop_over_expected <- o_win_prop |> 
  ggplot() +
  common_heatmap_layers() +
  pre_cumu_heatmap_tile(.fill = 'diff_win_series_prop') +
  add_text_layers(.f = add_o_series_win_over_expected_text_layer) +
  scale_fill_gradient2(
    low = extract_ggsci_mid_color('brown'),
    high = extract_ggsci_mid_color('cyan'),
    mid = 'white',
    midpoint = 0,
    limits = c(
      -max_diff_win_series_prop, 
      +max_diff_win_series_prop
    )
  ) +
  labs(
    title = 'Offensive Series Win % Over Expected'
  )
p_o_series_win_prop_over_expected
save_heatmap(p_o_series_win_prop_over_expected, width = 7.5, height = 5)
```

@@

```{r}
#| label: series_outcomes
prior_clinching_round_win_prop <- rounds |> 
  filter(round == (n_rounds - 1L)) |> 
  count(n_rounds, win_series, won_prior_round = win_round) |> 
  mutate(
    across(win_series, ~ifelse(.x, 'win', 'loss'))
  ) |> 
  pivot_wider(
    names_from = win_series,
    values_from = n,
    values_fill = 0L
  ) |> 
  filter(won_prior_round) |> 
  select(-won_prior_round) |> 
  mutate(
    win_series_prop = win / (win + loss),
    e_win_series_prop = (6L - 1L) / (n_rounds - 1L)
  )

n_rounds_max <- 11
n_sims <- 100000
set.seed(42)
w <- sample(c(0, 1), size = n_rounds_max * n_sims, replace = TRUE)

sim_rounds <- tibble(
  i = rep(1:n_sims, each = n_rounds_max),
  r = rep(1:n_sims, times = n_rounds_max),
  w = w
) |> 
  group_by(i) |> 
  mutate(
    cumu_w = cumsum(w),
    cumu_l = cumsum(w == 0)
  ) |> 
  ungroup() |> 
  mutate(
    cumu_wl_max = ifelse(cumu_w > cumu_l, cumu_w, cumu_l)
  ) |> 
  filter(cumu_wl_max <= 6)

sim_rounds_to_drop <- anti_join(
  sim_rounds |> 
    filter(cumu_wl_max == 6L),
  sim_rounds |> 
    filter(cumu_wl_max == 6L) |> 
    group_by(i) |> 
    slice_min(r, n = 1) |> 
    ungroup()
)

sim_rounds <- sim_rounds |> 
  anti_join(
    sim_rounds_to_drop
  )

## always from offensive perspective
e_series_streaks <- sim_rounds |> 
  inner_join(
    sim_rounds |> 
      filter(cumu_wl_max == 6) |> 
      mutate(
        zeros_win = cumu_l == cumu_wl_max
      ) |> 
      select(i, zeros_win)
  ) |> 
  mutate(
    across(w, ~ifelse(zeros_win, abs(1 - .x), .x))
  ) |> 
  group_by(i) |> 
  summarize(
    cumu_w = max(cumu_w),
    cumu_l = max(cumu_l),
    ws = paste0(w, collapse = '')
  ) |> 
  ungroup() |> 
  mutate(
    wins = ifelse(cumu_w == 6, cumu_w, cumu_l),
    losses = ifelse(cumu_w == 6, cumu_l, cumu_w)
  ) |> 
  unite(
    record, wins, losses, sep = '-'
  ) |> 
  count(record, ws, sort = TRUE) |> 
  mutate(prop = n / sum(n))

actual_series_streaks <- rounds |> 
  filter(win_series) |> 
  mutate(across(win_round, as.integer)) |> 
  group_by(year, event, series) |> 
  summarize(
    wins = max(cumu_w),
    losses = max(cumu_l),
    ws = paste0(win_round, collapse = '')
  ) |> 
  ungroup() |> 
  unite(
    record, wins, losses, sep = '-'
  ) |> 
  count(record, ws, sort = TRUE) |> 
  mutate(prop = n / sum(n))

series_streaks <- full_join(
  actual_series_streaks |> 
    rename_with(~sprintf('%s_actual', .x), c(n, prop)),
  e_series_streaks |> 
    rename_with(~sprintf('%s_expected', .x), c(n, prop))
) |> 
  mutate(
    prop_diff = prop_actual - prop_expected
  ) |> 
  arrange(desc(abs(prop_diff)))
series_streaks |> filter(is.na(n_actual))

## 6-0, 6-1, and some combos of 6-5 occur more often than expected
series_streaks |> 
  filter(!is.na(n_actual)) |> 
  mutate(total_actual = sum(n_actual)) |> 
  mutate(
    p = vectorized_prop_test(n_actual, total_actual, n_expected, !!n_sims)
  ) |> 
  select(-total_actual) |> 
  arrange(p$p.value)

## 6-0 and 6-1 occur more often than expected
series_outcomes <- full_join(
  actual_series_streaks |> 
    group_by(record) |> 
    summarize(
      across(n, sum)
    ) |> 
    mutate(prop = n / sum(n)) |> 
    rename_with(~sprintf('%s_actual', .x), c(n, prop)),
  e_series_streaks |> 
    group_by(record) |> 
    summarize(
      across(n, sum)
    ) |> 
    mutate(prop = n / sum(n)) |> 
    rename_with(~sprintf('%s_expected', .x), c(n, prop))
) |> 
  mutate(
    prop_diff = prop_actual - prop_expected
  ) |> 
  arrange(desc(abs(prop_diff))) |> 
  mutate(total_actual = sum(n_actual)) |> 
  mutate(
    p = vectorized_prop_test(n_actual, total_actual, n_expected, !!n_sims)
  ) |> 
  select(-total_actual) |> 
  arrange(p$p.value)

series_outcomes |> 
  select(record, actual = prop_actual, expected = prop_expected, diff = prop_diff) |> 
  pivot_longer(
    -record,
    names_to = 'prop',
    values_to = 'value'
  ) |> 
  mutate(
    across(
      prop,
      factor, 
      levels = c('actual', 'expected', 'diff')
    )
  ) |> 
  ggplot() +
  aes(x = record, y = prop) +
  geom_tile(
    show.legend = FALSE,
    aes(fill = value)
  ) +
  scale_x_discrete(
    position = 'top'
  )

```

@@

## When do back-to-back round wins occur?

```{r}
summarize_b2b_round_wins <- function(df, ...) {
  df |> 
    arrange(year, sheet, series, team, round) |> 
    group_by(year, sheet, series, team) |> 
    mutate(
      b2b_round_wins = win_round & lag(win_round)
    ) |> 
    ungroup() |> 
    filter(round > 1L) |> 
    count(..., round, b2b_round_wins) |> 
    group_by(..., round) |> 
    mutate(total = sum(n), prop = n / total) |> 
    ungroup() |> 
    filter(b2b_round_wins) |> 
    select(-b2b_round_wins)
}

pre_cumu_heatmap_tile <- function(..., .fill) {
  heatmap_tile(
    .xmin = 'n_rounds',
    .ymin = 'round',
    .fill = .fill,
    ...
  )
}

## now accounting for number of rounds that end up being played...
## seems like there is interesting behavior in the last round, regardless `n_rounds`
b2b_round_win_props <- rounds |> 
  filter(n_rounds > 6L) |> 
  summarize_b2b_round_wins(n_rounds)

b2b_round_win_props |> 
  ggplot() +
  pre_cumu_heatmap_tile(.fill = 'prop')

b2b_round_win_props |> 
  filter(n_rounds > 6L) |> 
  summarize_b2b_round_wins(n_rounds) |> 
  group_by(n_rounds) |> 
  mutate(
    intra_prop = (n - mean(n)) / sd(n)
  ) |> 
  ggplot() +
  aes(x = n_rounds, y = round) +
  geom_tile(aes(fill = intra_prop))

## now accounting for number of rounds that end up being played AND series winners...
rounds |> 
  filter(win_series, n_rounds > 6L) |> 
  summarize_b2b_round_wins(n_rounds)

## and the losers...
rounds |> 
  filter(!win_series) |> 
  summarize_b2b_round_wins(n_rounds)
```

@@

::: notes
-   Should you plant at most prevalent/winningest site? plant at other site? Go for kills?
:::
